Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-16 ind_files/01_Introduction.tex
Minor phrasing changes

\begin{itemize}
		\item Chapter \ref{chap_math_background} provides the necessary mathematical background and provides a toy example studying the size of a population of identical individuals that illustrates the major ideas used. If I am required to shoehorn the structure into the `intro-methods-results-discussion' format, then this chapter can be thought of as providing a mathematical `introduction' and illustration of the `methods' that will be used/generalized to get `results' in the following chapters.
		\item Chapter \ref{chap_BD} develops a formalism for describing the evolution of finite fluctuating populations of individuals that come in arbitrarily many `types' which vary in a \emph{discrete} character. This yields equations that generalize the classic Price equation and replicator-mutator equation to finite fluctuating populations.
		\item Chapter \ref{chap_infD_processes} extends the ideas developed in chapter \ref{chap_BD} to populations that vary in a single one-dimensional \emph{quantitative} character and derives a so-called `stochastic field theory' that describes evolution in such populations. This also results in some mathematical equations that may be of independent interest to physicists and applied mathematicians.
		\item Chapter \ref{chap_unification} provides a technical summary of the major results by deriving three important stochastic differential equations. These are equations for type frequencies, the population mean value of an arbitrary type-level quantity, and the population variance of an arbitrary type-level quantity, and respectively generalize the replicator-mutator equation, the Price equation, and \cite{lion_theoretical_2018}'s variance equation to finite, stochastically fluctuating populations. These equations predict a directional evolutionary force called `noise-induced selection' that is only seen in finite, fluctuating populations. Some implications for social evolution and community ecology are discussed. I also briefly discuss the field equation formalism I develop in chapter \ref{chap_infD_processes}.
		\item Chapter \ref{sec_disc} provides a quick summary of the major results and discusses biological implications, connections with previous studies, and opportunities for future work. This chapter has no equations (!). Thus, readers who do not want any mathematical equations can skip all other chapters and directly read chapter \ref{sec_disc} if they are interested only in the final results and takeaways\footnote{Though I promise I have tried my best to make sure the journey to this destination is also pleasant :)} 
\end{itemize}

>>>>

\begin{itemize}
		\item Chapter \ref{chap_math_background} provides the necessary mathematical background and provides a toy example studying the size of a population of identical individuals that illustrates the major ideas used. If I am required to shoehorn the structure into the `intro-methods-results-discussion' format, then this chapter can be thought of as providing a mathematical `introduction' and illustration of the `methods' that will be used/generalized to get `results' in the following chapters.
		\item Chapter \ref{chap_BD} develops a formalism for describing the evolution of finite fluctuating populations of individuals that come in arbitrarily many `types' which vary in a \emph{discrete} character. This yields equations that generalize the classic Price equation and replicator-mutator equation to finite fluctuating populations.
		\item Chapter \ref{chap_infD_processes} extends the ideas developed in chapter \ref{chap_BD} to populations that vary in a single one-dimensional \emph{quantitative} character and derives a so-called `stochastic field theory' that describes evolution in such populations. This also results in some mathematical equations that may be of independent interest to physicists and applied mathematicians.
		\item Chapter \ref{chap_unification} provides a technical summary of the major results by deriving three important stochastic differential equations. These are equations for type frequencies, the population mean value of an arbitrary type-level quantity, and the population variance of an arbitrary type-level quantity, and respectively generalize the replicator-mutator equation, the Price equation, and \cite{lion_theoretical_2018}'s variance equation to finite, stochastically fluctuating populations. These equations predict a directional evolutionary force called `noise-induced selection' that is only seen in finite, fluctuating populations. Some implications for social evolution and community ecology are discussed. I also briefly discuss the field equation formalism I develop in chapter \ref{chap_infD_processes}.
		\item Chapter \ref{sec_disc} provides a quick summary of the major results and discusses biological implications, connections with previous studies, and opportunities for future work. This chapter has no equations (!). Thus, readers who do not want any mathematical equations can skip all other chapters and directly read chapter \ref{sec_disc} if they are interested only in the final results and takeaways\footnote{Though I promise I have tried my best to make sure the journey to this destination is also pleasant :)}. 
\end{itemize}

### 2023-03-16 ind_files/06_nontechnical_discussion.tex
Minor phrasing changes

In Chapter \ref{chap_infD_processes}, I also postulate a `stochastic field theory' approach to modelling the evolution of quantitative traits in finite, fluctuating models. I show that this approach is consistent with known frameworks in theoretical population biology, reproducing results from quantitative genetics in the infinite population limit. The field equations I formulate are exactly complementary to a recent approach that uses spacetime white noise~\citep{week_white_2021}, and the connections are discussed in detail in section \ref{sec_disc_field_eqns}. This formulation also highlights a (somewhat heuristic) approach to the study of spacetime stochastic processes and related mathematical objects that avoids measure-theoretic tools and may be of independent interest to applied mathematicians. Broadly similar stochastic field theoretic approaches have also been proposed in mathematical neurobiology~\citep{buice_field-theoretic_2007,bressloff_stochastic_2010,coombes_neural_2014} and models of collective motion~\citep{o_laighleis_minimal_2018}. I  conjecture that the stochastic `fundamental equations' I derive in section \ref{sec_fun_theorems} for discrete traits should also hold for quantitative traits under mild assumptions on the trait space of the quantitative traits, though no proof is attempted in this thesis. We can, however, take encouragement from the observation that~\cite{week_white_2021} have already derived some special cases of these equations\footnote{In particular, equations (21b) and (21c) in \cite{week_white_2021} are precisely the $m \to \infty$ limit of my equations for changes in the mean value of a type-level quantity (equation \ref{nD_stochastic_Price}) and changes in the variance of a type-level quantity (equation \ref{nD_stochastic_Price_variance}) respectively for the special case in which the type-level quantity is the value of the quantitative trait being studied. See section \ref{sec_disc_field_eqns} for a more detailed technical discussion.} for quantitative traits using a measure-theoretic approach under certain technical assumptions on the stochastic process under study, the most biologically important assumption being that the traits are normally distributed in the population. My formalism and \cite{week_white_2021}'s formalism are complementary to each other, and reflect a deep duality between related mathematical structures that is well-appreciated in the applied mathematics literature. These two complementary approaches to modeling stochastic processes have long been recognized in population genetics, where the (simpler analogs of the) general approach in models of fixed total population size go by the names `branching processes approach'~(\cite{week_white_2021}'s formalism; In the classic literature, seen in work like~\cite{haldane_mathematical_1927,fisher_distribution_1931}) and `diffusion theory approach'~(My formalism; In the classic literature, seen in work like~\cite{fisher_dominance_1923,wright_evolution_1931,kimura_problems_1957}).

>>>>

In Chapter \ref{chap_infD_processes}, I also postulate a `stochastic field theory' approach to modelling the evolution of quantitative traits in finite, fluctuating models. I show that this approach is consistent with known frameworks in theoretical population biology, reproducing results from quantitative genetics in the infinite population limit. This formulation also highlights a (somewhat heuristic) approach to the study of spacetime stochastic processes and related mathematical objects that avoids measure-theoretic tools and may be of independent interest to applied mathematicians. Broadly similar stochastic field theoretic approaches have also been proposed in mathematical neurobiology~\citep{buice_field-theoretic_2007,bressloff_stochastic_2010,coombes_neural_2014} and models of collective motion~\citep{o_laighleis_minimal_2018}. I  conjecture that the stochastic `fundamental equations' I derive in section \ref{sec_fun_theorems} for discrete traits should also hold for quantitative traits under mild assumptions on the trait space of the quantitative traits, though no proof is attempted in this thesis. We can, however, take encouragement from the observation that~\cite{week_white_2021} have already derived some special cases of these equations\footnote{In particular, equations (21b) and (21c) in \cite{week_white_2021} are precisely the $m \to \infty$ limit of my equations for changes in the mean value of a type-level quantity (equation \ref{nD_stochastic_Price}) and changes in the variance of a type-level quantity (equation \ref{nD_stochastic_Price_variance}) respectively for the special case in which the type-level quantity is the value of the quantitative trait being studied. See section \ref{sec_disc_field_eqns} for a more detailed technical discussion.} for quantitative traits using an approach grounded in the theory of measure-valued branching processes under certain technical assumptions on the stochastic process under study, the most biologically important assumption being that the traits are normally distributed in the population. My formalism and \cite{week_white_2021}'s formalism are complementary to each other, and reflect a deep duality between related mathematical structures that is well-appreciated in the applied mathematics literature (see section \ref{sec_disc_field_eqns}). These two complementary approaches to modeling stochastic processes have long been recognized in population genetics, where the (simpler analogs of the) general approach in models of fixed total population size go by the names `branching processes approach'~(\cite{week_white_2021}'s formalism; In the classic literature, seen in work like~\cite{haldane_mathematical_1927,fisher_distribution_1931}) and `diffusion theory approach'~(My formalism; In the classic literature, seen in work like~\cite{fisher_dominance_1923,wright_evolution_1931,kimura_problems_1957}).
